What We Eval in the Shadows

### 2021-08-11 paper/main.tex

\begin{tabular}{@{}p{.97\linewidth}}
  \medskip\EE{$min(e)=\c{V}$} \\[-2mm]\small Values occur in 16\% of sites, 74\% of
  those are constants such as integers, the remainder evaluate to a value. \Eval
  is needed when the expression denoting the values is constructed
  programmatically. Cases when a value is directly given to \eval, and \eval
  returns it unchanged, often occur when a computation has a default path and
  another, more interesting, path that requires evaluation. The following is a
  call site that always see a string; \packageNbCallSitesUniqueActualValue of
  sites only ever see a value. This shape usually run in few interpreter steps.
  The environment in which they evaluate is mostly irrelevant.
\begin{lstlisting}
  eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
\\
\medskip\EE{$min(e)=\c{X}$}\\[-2mm]\small Variable lookups are the most common shape,
\packageNbSymbolVarSitePercent are simple variable names, e.g. \c x. The shapes
subsumes \c{V} and built-in arithmetic operations, so a mixture of arithmetics
and variables normalize to \c X. A lookup is one step of execution, if a promise
is returned an arbitrary number of additional steps may be needed; the median is
\packageMinimizedmedianoperationsaRnd, suggesting that most variable read do
little work. Variables are often evaluated in environments that have been
constructed programmatically, \packageMinimizedpercentparentframesa of
expressions are evaluated in a function environment.
\\
\medskip\EE{$min(e)=\c{\$}$}\\[-2mm]\small This shape extends \c X to include lookup
with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} and
\c{x[[24]]}. This takes a few more step of evaluation on average, and is
typically used in a function environment.
\\
\medskip\EE{$min(e)=$~\c{<-}}\\[-2mm]\small This includes both direct assignment {\tt
  <-} and parent environment assignment {\tt <\,\!<-} as well as the \c{assign}
function. It subsumes \c{\$}. Assignments
represent the most obvious source of side effects in \eval.
\\
\medskip\EE{$min(e)=\c{F()}$}\\[-2mm]\small Function calls without variable lookups or
assignment. A variable is allowed the function position, looking up
functions does not trigger computation unless the function's name
is bound to a promise. This shape typically does not perform much work.
\\
\medskip\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small Function
calls whose arguments may include variable references and nested calls. Together
they represent the most frequent shape and perform over 16 steps. They are
frequently evaluated in function environments.
\end{tabular}

@@

Out of the minimized expressions, it is \c{BLOCK}, \c{<-} and \c{F(F(X)} that
contribute to the vast majority of side effects. Concretely, 92\% of \c{BLOCK},
87\% of \c{<-} and 15\% of \c{F(F(X))} expressions do side effects. In the case
of \c{BLOCK}, most of them happen in the environment passed to the \eval call
(default is the parent environment). For \c{<-} it is 15\%. However, this form
is dominated by the \c{plyr::allocate\_column} function that side-effects in
the environment containing the data frame to which it allocates new column.
Without this call, \c{<-} is similarly predictable with 96\% of side effects
happening in the passed environment. For \c{F(F(X))}, it is less clear as 58\%
of side effects happen in a different environment to the one that was passed to
\eval.

\paragraph{Discussion} In JavaScript assignments in \eval can happen in either
local scope or less often, when called through an alias, in the global scope. In
R, given the support for first-class environments, it can happen anywhere,
making it \eval more dangerous than it already is. However, the data suggest
that first, side-effects from \eval are not as widespread as in the case of
JavaScript,\footnote{\citep{ecoop11} shows that in the \emph{Interactive}
scenario, \eval in Javascript performs, store events can reach up to 40\% of the
events, and 7\% to 8\% of the \eval do side effects in the global scope. } and
that over half of them happen in a predictable environment.

@@

>>>>

\begin{tabular}{@{}p{.97\linewidth}}
  \medskip\EE{$min(e)=\c{V}$}\\[-2mm]\small Values occur in 16\% of sites, 74\% of
  those are constants such as integers, the remainder evaluate to a value (\packageNbCallSitesUniqueActualValue of
  sites only ever see a value). \Eval
  is needed when the expression denoting the values is constructed
  programmatically. Cases when a value is directly given to \eval, and \eval
  returns it unchanged, often occur when a computation has a default path and
  another, more interesting, path that requires evaluation.
  This shape usually run in few interpreter steps.
  The environment in which they evaluate is mostly irrelevant.
\begin{lstlisting}
  eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
\\
\medskip\EE{$min(e)=\c{X}$}\\[-2mm]\small Variable lookups are the most common shape,
\packageNbSymbolVarSitePercent are simple variable names, e.g. \c x. The shapes
subsumes \c{V} and built-in arithmetic operations, so a mixture of arithmetics
and variables normalize to \c X. A lookup is one step of execution, if a promise
is returned an arbitrary number of additional steps may be needed; the median is
\packageMinimizedmedianoperationsaRnd, suggesting that most variable read do
little work. Variables are often evaluated in environments that have been
constructed programmatically, \packageMinimizedpercentparentframesa of
expressions are evaluated in a function environment.
\\
\medskip\EE{$min(e)=\c{\$}$}\\[-2mm]\small This shape extends \c X to include lookup
with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} and
\c{x[[24]]}. This takes a few more step of evaluation on average, and is
typically used in a function environment.
\\
\medskip\EE{$min(e)=$~\c{<-}}\\[-2mm]\small This includes both direct assignment {\tt
  <-} and parent environment assignment {\tt <\,\!<-} as well as the \c{assign}
function. It subsumes \c{\$}. Assignments
represent the most obvious source of side effects in \eval.
\\
\medskip\EE{$min(e)=\c{F()}$}\\[-2mm]\small Function calls without variable lookups or
assignment. A variable is allowed the function position, looking up
functions does not trigger computation unless the function's name
is bound to a promise. This shape typically does not perform much work.
\\
\medskip\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small Function
calls whose arguments may include variable references and nested calls. Together
they represent the most frequent shape and perform over 16 steps. They are
frequently evaluated in function environments.
\end{tabular}

@@

Out of the minimized expressions, it is \c{BLOCK}, \c{<-} and \c{F(F(X))} that
contribute to the vast majority of side effects. Concretely, 92\% of \c{BLOCK},
87\% of \c{<-} and 15\% of \c{F(F(X))} expressions do side effects. In the case
of \c{BLOCK}, most of them happen in the environment passed to the \eval call
(default is the parent environment). For \c{<-} it is 15\%. However, this form
is dominated by the \c{plyr::allocate\_column} function that side-effects in
the environment containing the data frame to which it allocates new column.
Without this call, \c{<-} is similarly predictable with 96\% of side effects
happening in the passed environment. For \c{F(F(X))}, it is less clear as 58\%
of side effects happen in a different environment to the one that was passed to
\eval.

\paragraph{Discussion} In JavaScript assignments in \eval can happen in either
local scope or less often, when called through an alias, in the global scope.
In R, given the support for first-class environments, it can happen anywhere,
making it \eval more dangerous than it already is. However, the data suggest
that first, side-effects from \eval are not as widespread as in the case of
JavaScript,\footnote{\citep{ecoop11} shows that in the \emph{Interactive}
  scenario, \eval in Javascript performs, store events can reach up to 40\% of
the events, and 7\% to 8\% of the \eval do side effects in the global scope. }
and that over half of them happen in a predictable environment. This gives a
ray of hope for a hypothetical R compiler. Even though \eval can do anything
anywhere, the data suggests most effects are sane.

@@

The use of \eval with \code{substitute} is a general pattern that we often see
in R packages. It is used to make the interfaces more flexible. For example the
\code{subset}, which selects columns in a data frame, uses it to allow to specify
the selected columns using either the column name or index. For example
\code{subset(df, sel=c(1,B,x))} returns a subset of \code{df} with the first
column, the column named \code{B} and whatever column is at index the variable
\code{x} holds.

It is implemented using \code{eval(substitute(sel), nl, parent.frame())}. The
first argument returns the expression passed to \code{sel}. The second is a
named list that contains the names of the columns from \code{df} and which will
be used to evaluate the expression. The last parameter indicates which
environment should be used to look up bindings not found in \c{nl}.
